Appendix S1. Linear regression models and non-linear population dynamics

### 2024-05-04 Appendix_S1_v01.Rmd
Begin 2-species competition + lm

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

```{r}
knitr::knit_exit()
```

# Simulate initial species population growth
N1.0 <- rpois(1,10)
r1.0 <- 1.67
alpha.11 <- 0.00125

equation becomes:

$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t}}$$

We make this not linked to then linked to growth and in each instance run a linear model to show no effect is detected or is detected as it should be. E can vary randomly, as long as it links to growth.

```{r}
# Simulate initial species population growth
N1.0 <- rpois(1,10)
r1.0 <- 1.67
alpha.11 <- 0.00125
# model function
disc_log <- function(r, N0, alpha) {
    Nt1 <- (r*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 30
N <- rep(NA, t)
N[1] <- N1.0
for (i in 2:t) {
  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
}
```

In the metacommunity simulation in the main text, a species $i$ resides in a site $k$ with an environmental optimum trait value $E_k \sim unif(0,1)$, an initial population size $N_{i,0} \sim Pois(10)$, and an initial degree of maladaptation $\beta_{0,i} \sim Gamma(0.75,1)$. From this \(\beta_{0,i}\), the initial distance to the local patch's optimum phenotype was calculated as $d_{0,i}=\sqrt{B_{0,i}(w+P)}$ and the initial trait value for each species $x_{0,i}$ is calculated by subtracting $d_{0,i}$ from the local environmental optimum value $E_k$. Here I simulate a species that is initially suited for the environment ($d_{0,i}$)

Internal question 1 - how does the $\alpha$ version compared to the $K$ version?

```{r}
set.seed(42)
# Simulate initial species population growth
N1.0 <- rpois(1,10)
# Behind-the-scenes initial r1.0
E1 <- runif(1,0,1)
beta1.0 <- rgamma(1,0.75,1)
w <- 10
P <- 1
Wmax <- 2
What <- Wmax*sqrt(w / (P+w))
h2 <- 0
d1.0 <- sqrt(beta1.0*(w+P))
x1.0 <- E1 - d1.0
# Emergent initial r1.0
r1.0 <- What*exp( -(((w+(1-h2)*P)/(P+w))*(E1-x1.0))^2/ (2*(P+w)))
alpha.11 <- 0.00125

# model function
disc_log <- function(r, N0, alpha) {
    Nt1 <- (r*N0) / (1+alpha*N0)
    return(Nt1)
}

# Simulation of model for t time steps
N <- rep(NA, t)
N[1] <- N1.0
t <- 30
for (i in 2:t) {
  N[i] <- disc_log(r=r1.0, N0=N[i-1], alpha=alpha.11)
    N0 <- N[i]
}
# Plot simulation: ggplot
dat <- as.data.frame(N)
dat$time <- as.numeric(rownames(dat))
ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = (1/alpha.11),
    linetype = "dashed", color = "gray")
# Additional plot: how does growth rate change with time?
range <- seq(0,800,by=0.01)
rec <- rep(NA, length(range))
for (i in 1:length(range)) {
    rec[i] <- r1.0 / (1+alpha.11*range[i])
}
plot(y=rec,x=range,ylim=c(-1,2))
```

```{r}
# Parameter values to use for simulation
r <- 1.21
K <- 1400
N0 <- 4
t <- 30
# model function
disc_log <- function(r, N0, K) {
    Nt1 <- N0 + r * N0 * (1 - N0/K)
    return(Nt1)
}
# Simulation of model for t time steps
N <- rep(NA, t)
for (i in 1:t) {
    N[i] <- disc_log(r, N0, K)
    N0 <- N[i]
}
# Plot simulation: ggplot
dat <- as.data.frame(N)
dat$time <- as.numeric(rownames(dat))
ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = K,
    linetype = "dashed", color = "gray")
```

$$ N_{i,t+1} =\frac{\hat{W}e^ \frac{-[(\frac{w+(1-h^2)P} {P+w})(E-x_{i,t})]^2}{2(P+w)}N_{i,t}} {1 + \alpha_{ii}N_{i,t} + \sum_{j \neq i}^{S} \alpha_{ij}N_{j,t}} $$

```{r}

```

## Generating the data frames
For the diagnosis first need to convert the chains into a coda object and extract the effective sample size (ESS), which denotes how many MCMC iteration were independent (e.g. having no autocorrelation) from previous iterations, and  potential scale reduction factors (PSRF), which describe how well the MCMC chains converged around a value and if the two chains differ from each other. Therefore a high value for ESS (close to the number of actual iterations multiplied by the number of chains) is preferable and a PSRF value of 1 represents both chains converging on the same mean.
Note we do not check the predictive and explanatory power of the model, as we have a perfect data set by the time series data of the simulations, which leads to $R^2$ > .98.
```{r eval=FALSE}
load("../../output/h_01_d_minus3_hmsc_v01.RData")
m.post <- convertToCodaObject(m.1)
ess.beta<- effectiveSize(m.post$Beta)
psrf.beta <- gelman.diag(m.post$Beta, multivariate=FALSE)$psrf
```

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

# Two species, logistic growth, competition
XXXXX
## Interspecific competition
The growth equation for each species now becomes:

$$ N_{i,t+1} = \frac{r_i N_{i,t}}{1 + \alpha_{ii} N_{i,t} + \alpha_{ij} N_{j,t}}$$
We use a distinct growth rate for the 2nd species and introduce the interspecific interaction coefficient $\alpha_{ij}$.

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
N2.0 <- 10
r1.0 <- 1.67
r2.0 <- 1.7
alpha.11 <- 0.00125
alpha.22 <- 0.00125
alpha.12 <- 0.008
alpha.21 <- 0.008725
# model function
disc_LV_comp <- function(r1,r2,N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21) {
    Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
    Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
    return(c(Nt1,Nt2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
for (i in 2:t) {
  res <- disc_LV_comp(r1=r1.0,r2=r2.0,N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
}
```
```{r figs11, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125. Relationship between E and Nt is also shown."}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
```
## Linear statistical model with environmental covariate
XX
## Bayesian linear statistical model: HMSC
XX
## Conclusions
XX
